Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-04-18 doc/manuscript.Rmd
Most edits done. Only a couple of small things.

**Word Count**: 238 / 250

@@

Most previous work has relied on studying bacterial and phage communities in isolation, reducing them to two separate communities. In reality, bacteria and phage communities are dynamic and complex. They frequently share genetic information and work together to maintain stable ecosystems. Removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. The most common approach of studying communities in isolation has been unable to capture how these communities interact and influence one another, but relationship-based network approaches allow us to capture this information. We aimed to bridge this knowledge gap by leveraging machine learning and graph theory techniques to characterize human bacterial and phage communities by their relationships. We focused on characterizing the network diversity of the human skin and gut microbiomes to provide a foundation for further studies of disease network dynamics and to gain broader insights into human microbiome network diversity across different body sites.

@@

## Cohort Curation and Sample Processing
We studied the differences in microbiome phage-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were all subjected to virus-like particle (VLP) purification to eliminate other organismal DNA and allow for direct assessment of both the extracellular and actively replicating intracellular virome.

We utilized data that was published over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. The bacterial and viral sequences from those publications were quality filtered and assembled into contigs. We generate clusters of related related bacterial and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm. We interpreted these clusters as operational units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht]. The resulting contigs and OGUs demonstrated high sequence coverage and length **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**.

## Establishing a Model of Phage-Bacteria Interactions
We predicted which phage OGUs infected which bacteria using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. This training set contained 43 diverse bacterial species and 30 diverse phage strains, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because this was a DNA sequencing study, RNA phages were not considered **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (phage infects a bacterium) and negative relationships (phage does not infect a bacterium). This allowed us to validate the false positive and false negative rates associated with our candidate models, thereby building upon previous work that only considered positive relationship [@Edwards:2015iz].

Four phage and bacterial genomic features were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families associated with experimentally identified protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were not informative important **(Figure \ref{RocCurve} B)**. Approximately one third of the relationships yielded no score and were therefore unknown, unable to be assigned an interaction prediction **(Figure \ref{RocCurve} C)**.

Next we used our random forest model to classify the relationships between bacteria and phage operational genomic units in the experimental datasets. The relationships predicted within the three published datasets were used to construct one master network, which was built to contain the three studies as sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and OGU abundance within the community were stored in the master network for downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` bacteria. Although the network was highly connected, not all relationships were present in all samples. As relationships were weighted by the relative abundances of their associated bacteria and phages, lowly abundant relationships could be present but not highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included almost all of the master network nodes and edges **(Figure \ref{RocCurve} E - F)**. The phages and bacteria in the gut diet and twin sample sets were more sparsely related, each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, which indicated more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

@@

We used this network dissimilarity metric to test whether microbiome network structures were more similar within people than between people over time. We found that gut microbiome network structure clustered by person (ANOSIM p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, R = `r signif(interstats[c(interstats$site %in% "AS"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} A**). Network dissimilarity within people over the 8-10 day sampling period was less than the average dissimilarity between that person and others, although this was not statistically significant (p-value = `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} B**). The lack of statistical confidence was likely due to the small gut microbiome sample size of this dataset. Although there was evidence for gut network conservation among individuals, we found no evidence for conservation of gut network structures within families. The gut network structures were not more similar between twins and their mothers (intrafamily) compared to other families of twins and mothers (inter-family) (p-value = `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} C**).

<!-- Maybe add twin to twin data here? -->

@@

## Role of Diet & Obesity in Gut Microbiome Conectivity

<!-- Could maybe move this section up before the previous section -->

Diet is a major environmental factor that influences resource availability and gut microbiome composition and diversity, including bacteria and phages [@Minot:2011ez; @Turnbaugh:2009hf; @David:2014cl]. Previous work in isolated culture-based systems has suggested that changes in nutrient availability are associated with altered phage - bacteria network structures [@Poisot:2011jc], although this has yet to be tested in humans. We hypothesized that a change in diet would be associated with a change in network structure in these human systems.

We evaluated the diet-associated differences in gut network structure by quantifying how central each sample's network was on average. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. Degree centrality, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. We more completely quantified the network properties by supplementing degree centrality with measurements of closeness centrality. Closeness centrality is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that the effects of genetic information or altered abundance would be more impactful to all other microbes in the system. A network with higher average closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience against instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

We found that the gut microbiome network structures associated with high-fat diets were less connected than those of low-fat diets **(Figure \ref{dietnetworks} A-B)**. Tests for statistical differences were not performed due to the small sample size. High-fat diets exhibited less degree centrality **(Figure \ref{dietnetworks} A)**. This finding suggests that bacteria in high-fat environments were targeted by fewer phages and that phage tropism was more specific. High-fat diets also exhibited decreased closeness centrality **(Figure \ref{dietnetworks} B)**, indicating that bacteria and phages were more distant from other bacteria and phages in the community. This would make genetic transfer and altered abundance of a given phage or bacterium less capable of impacting other bacteria and phages within the network.

@@

We found that, just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were also conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggest that microbiome network properties, such as stability and the potential for horizontal gene transfer, are personal and may be impacted by personal factors ranging from the body's immune system to environmental conditions such as climate and diet. The ability of environmental conditions to shape gut and skin microbiome network structure was supported by our finding that diet and skin location were associated with altered network structures.

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, we found evidence that high-fat diets were less connected than low-fat diets. This suggested that high-fat diets may lead to less stable communities with a decreased ability for individuals to influence one other. We supported this finding with the observation that obesity was associated with decreased network connectivity. Together these findings suggest the food we eat may not only impact which microbes colonize our guts, but may also impact their interactions. Further work will be required to characterize these relationships with a larger cohort.

@@

![**Summary information of validation dataset used in the interaction predictive model.** *A) Categorical heat-map highlighting the experimentally validated positive and negative interactions. Only bacteria species are shown, which represent multiple reference strains. Phages are labeled on the x-axis and bacteria are labeled on the y-axis. B) World map illustrating the sampling locations used in the study (red dots). C) Quantification of bacterial host strains known to exist for each phage. D) Genome strandedness and E) linearity of the phage reference genomes used for the dataset.*\label{ValidationOverview}](../figures/BenchmarkDataset.pdf){ width=100% }

>>>>

**Word Count**: 237 / 250

@@

Most previous work has relied on studying bacterial and phage communities in isolation, reducing them to two separate communities [@Norman:2015kb; @Manrique:2016dx; @Ly:2014ew; @Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. In reality, bacteria and phage communities are dynamic and complex. They frequently share genetic information and work together to maintain stable ecosystems. Removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. The most common approach to studying communities in isolation has been unable to capture how these communities interact and influence one another, but relationship-based network approaches allow us to capture this information. We aimed to bridge this knowledge gap by leveraging machine learning and graph theory techniques to characterize human bacterial and phage communities by their relationships. We focused on characterizing the human skin and gut virome-microbiome network diversity to provide a foundation for further studies of disease network dynamics and to gain broader insights into human microbiome network diversity across different body sites.

@@

## Cohort Curation and Sample Processing
We studied the differences in virome-microbiome interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were selected because they were subjected to virus-like particle (VLP) purification to eliminate other organismal DNA and allow for direct assessment of both the extracellular and actively replicating intracellular virome.

We utilized data that was published over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. The bacterial and viral sequences from those publications were quality filtered and assembled into contigs. We generated clusters of related related bacterial and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. We interpreted these clusters as operational units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht].

## Establishing a Model of Phage-Bacteria Interactions
We predicted which phage OGUs infected which bacterial OGUs using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. This training set contained 43 diverse bacterial species and 30 diverse phage strains, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because we used DNA sequencing studies, RNA phages were not considered **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (phage infects a bacterium) and negative relationships (phage does not infect a bacterium). This allowed us to validate the false positive and false negative rates associated with our candidate models, thereby building upon previous work that only considered positive relationships [@Edwards:2015iz].

Four phage and bacterial genomic features were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families associated with experimentally identified protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were not informative **(Figure \ref{RocCurve} B)**. Approximately one third of the relationships yielded no score and were therefore unable to be assigned an interaction prediction **(Figure \ref{RocCurve} C)**.

Once our random forest model was built, we used it to classify the relationships between bacteria and phage operational genomic units in the experimental datasets. The relationships predicted within the three published datasets were used to construct one master network, which was built to contain the three studies as sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and OGU abundance within the community were stored in the master network for downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` bacteria. Although the network was highly connected, not all relationships were present in all samples. As relationships were weighted by the relative abundances of their associated bacteria and phages, lowly abundant relationships could be present but not highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Vertices"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], digits = 1)`% and `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Edges"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], digits = 1)`% of the master network nodes and edges, respectively **(Figure \ref{RocCurve} E - F)**. The phages and bacteria in the gut diet and twin sample sets were more sparsely related, each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, which indicated more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

## Role of Diet & Obesity in Gut Microbiome Connectivity

Diet is a major environmental factor that influences resource availability and gut microbiome composition and diversity, including bacteria and phages [@Minot:2011ez; @Turnbaugh:2009hf; @David:2014cl]. Previous work in isolated culture-based systems has suggested that changes in nutrient availability are associated with altered phage - bacteria network structures [@Poisot:2011jc], although this has yet to be tested in humans. We hypothesized that a change in diet would be associated with a change in virome-microbiome network structure in these human systems.

We evaluated the diet-associated differences in gut virome-microbiome network structure by quantifying how central each sample's network was on average. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. Degree centrality, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. We also quantified the network properties by supplementing degree centrality with measurements of closeness centrality. Closeness centrality is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that the effects of genetic information or altered abundance would be more impactful to all other microbes in the system. A network with higher average closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience against instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

We found that the gut microbiome network structures associated with high-fat diets were less connected than those of low-fat diets **(Figure \ref{dietnetworks} A-B)**. Tests for statistical differences were not performed due to the small sample size. High-fat diets exhibited less degree centrality **(Figure \ref{dietnetworks} A)**. This finding suggested that bacteria in high-fat environments were targeted by fewer phages and that phage tropism was more specific. High-fat diets also exhibited decreased closeness centrality **(Figure \ref{dietnetworks} B)**, indicating that bacteria and phages were more distant from other bacteria and phages in the community. This would make genetic transfer and altered abundance of a given phage or bacterium less capable of impacting other bacteria and phages within the network.

@@

We used this network dissimilarity metric to test whether microbiome network structures were more similar within people than between people over time. We found that gut microbiome network structure clustered by person (ANOSIM p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, R = `r signif(interstats[c(interstats$site %in% "AS"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} A**). Network dissimilarity within each person over the 8-10 day sampling period was less than the average dissimilarity between that person and others, although this difference was not statistically significant (p-value = `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} B**). The lack of statistical confidence was likely due to the small sample size of this dataset. Although there was evidence for gut network conservation among individuals, we found no evidence for conservation of gut network structures within families. The gut network structures were not more similar within families (twins and their mothers; intrafamily) compared to other families (inter-family) (p-value = `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} C**).

@@

@@

We found that, just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were also conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggested that microbiome network properties, such as stability and the potential for horizontal gene transfer, were personal and may be impacted by personal factors ranging from the body's immune system to environmental conditions such as climate and diet. The ability of environmental conditions to shape gut and skin microbiome network structure was supported by our finding that diet and skin location were associated with altered network structures.

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, we found evidence that high-fat diets were less connected than low-fat diets. This suggested that high-fat diets may lead to less stable communities with a decreased ability for individuals to influence one other. We supported this finding with the observation that obesity was associated with decreased network connectivity (closeness of bacterial and phage vertices). Together these findings suggest the food we eat may not only impact which microbes colonize our guts, but may also impact their interactions. Further work will be required to characterize these relationships with a larger cohort.

@@

![**Summary information of validation dataset used in the interaction predictive model.** *A) Categorical heat-map highlighting the experimentally validated positive and negative interactions. Only bacteria species are shown, which represent multiple reference strains. Phages are labeled on the x-axis and bacteria are labeled on the y-axis. B) Quantification of bacterial host strains known to exist for each phage. C) Genome strandedness and E) linearity of the phage reference genomes used for the dataset.*\label{ValidationOverview}](../figures/BenchmarkDataset.pdf){ width=100% }
